---
title: 'Lab 3: Dependence -- solution'
---

```{r, include = FALSE}
# This option turns evaluation of R code off in this document. Remove it
# if you use it as a template for solving the lab exercise.
knitr::opts_chunk$set(eval = TRUE)
```

This session covers tools for investigating
dependence between points.

```{r, message=FALSE}
library(spatstat)
```

### Exercise 1

The `swedishpines` dataset was recorded in a study plot
in a large forest. We shall assume the pattern is stationary.

1.  Calculate the estimate of the $K$-function using `Kest`.

    The estimation is done with:
    
    ```{r}
    K <- Kest(swedishpines)
    ```

2.  Plot the estimate of $K(r)$ against $r$

    To plot the K-function, we do:
    
    ```{r, results='hide'}
    plot(K, main = "K-function")
    ```

3.  Plot the estimate of $K(r) - \pi\!r^2$ against $r$ (Hint: look at the `fmla`
    argument in `plot.fv`).

    The estimated K-function subtracted $\pi\!r^2$ can be done via the `fmla`
    (formula) interface:

    ```{r}
    plot(K, . - pi*r^2 ~ r, main = "Normalized K-function",
         legendpos = "bottomright")
    ```
    
4.  Calculate and plot an estimate of the pair correlation function
    using `pcf`.
  
    The pair-correlation is also compute straight-forwardly:
    
    ```{r}
    pcorf <- pcf(swedishpines)
    plot(pcorf)
    ```

5.  Draw tentative conclusions from these plots about interpoint
    interaction in the data.

    Assuming a homogeneous point pattern, both the L- and K-function are less
    what is expected under the Poisson process the data. Thus they indicate a
    comparatively regular point pattern. Similarly, the pair-correlation
    function also suggests this.

### Exercise 2

The command `rThomas` generates simulated realisations of
the Thomas model (‘modified Thomas cluster process’).

1.  Read the help file.

    See `help("rThomas")`.

2.  Type `plot(rThomas(10, 0.05, 8))` a few times, and
    interpret the results.

    ```{r results="hide"}
    replicate(3, plot(rThomas(10, 0.05, 8), main = ""))
    ```

    A clustered process -- on average 10 clusters with 8 points. The standard
    deviation of the cluster distribution is 0.05, so most points will be within
    distance 0.10 from their parent.

3.  Experiment with the arguments of `rThomas` to obtain
    point patterns that
    
    i.  consist of a few, well-separated, very tight clusters of
        points;
        
    ii. look similar to realisations of a uniform Poisson process.

    We get few clusters by reducing the intensity of the parent process (first argument).
    Tightly and separated clusters are obtained by reducing the standard deviation
    (second argument).

    ```{r}
    plot(rThomas(5, scale = 0.01, mu = 8), main = "")
    ```

    If the are many clusters with a large standard deviation it looks like Poisson.

    ```{r}
    plot(rThomas(100, scale = 1, mu = 1), main = "")
    ```

### Exercise 3

1.  Read the help file for `kppm`.

    See `help("kppm")`.

2.  Fit the Thomas model to the `redwood` data by the
    method of minimum contrast:
    ```{r, eval=FALSE}
    fit <- kppm(redwood ~ 1, clusters="Thomas")
    fit
    plot(fit)
    ```

    From the documentation, the minmum contrast fitting procedure is default.
    Hence, we need not specify it.

    ```{r}
    fit <- kppm(redwood ~ 1, clusters = "Thomas")
    fit
    plot(fit, main = "", pause = FALSE) 
    ```

3.  Read off the parameters of the fitted model, and generate a
    simulated realisation of the fitted model using `rThomas`.

    From the previous output, we can read off the parameters to do the simulation
    (or we can use `parameters` to extract them):

    ```{r}
    (p <- parameters(fit))
    rt2 <- rThomas(kappa = p$kappa, scale = p$scale, mu = p$mu)
    plot(rt2, main = "")
    ```

4.  Type `plot(simulate(fit))` to generate a simulated realisation
    of the fitted model automatically.

    OK, let try that alternative:
    
    ```{r}
    plot(simulate(fit, drop = TRUE), main = "")
    ```

5.  Try the command
    ```{r eval=FALSE}
    fit2 <- kppm(redwood ~ 1, clusters="Thomas", startpar=c(kappa=10, scale=0.1))
    ```
    and briefly explore the fitting algorithm’s sensitivity to the
    initial guesses at the parameter values `kappa` and `scale`.

    For "large" kappa (parent intensity) and "small" scale (standard deviation), the 
    algorithm seems quite robust:
    ```{r}
    kppm(redwood ~ 1, clusters="Thomas", startpar=c(kappa=10, scale=0.1))
    kppm(redwood ~ 1, clusters="Thomas", startpar=c(kappa=100, scale=0.01))
    ```

    However, for a very small parent intensity (kappa) and large offspring scale the
    fit changes considerably.
    
    ```{r}
    kppm(redwood ~ 1, clusters="Thomas", startpar=c(kappa=0.1, scale=10))
    ```

6.  Generate and plot several simulated realisations of the fitted
    model, to assess whether it is plausible.

    ```{r}
    XX <- simulate(fit, nsim = 11)
    XX[[12]] <- redwood
    plot(XX, main = "", main.panel = "")
    ```
    
    The actual data do not look too different from the simulated (apart from the
    artificial discretisation in the real data which can be seen on larger plots).

7.  Extract and plot the fitted pair correlation function by
    ```{r, eval = FALSE}
    pcffit <- pcfmodel(fit)
    plot(pcffit, xlim = c(0, 0.3))
    ```

    OK, let's try that:
    
    ```{r}
    pcffit <- pcfmodel(fit)
    plot(pcffit, xlim = c(0, 0.3), main = "pair correlation")
    ```

8.  Type `plot(envelope(fit, Lest, nsim=39))` to generate simulation
    envelopes of the $L$ function from this fitted model. Do they
    suggest the model is plausible?

    ```{r}
    plot(envelope(fit, Lest, nsim = 39, global = TRUE))
    ```
    
    Yes, the model seems plausible and it does not devivate from the envelope.


### Exercise 4

1.  Fit a Matern cluster process to the `redwood` data.

    We fit the Matern cluster process by specifying the `clusters` argument to
    be `MatClust`.
    
    ```{r}
    mfit <- kppm(redwood ~ 1, clusters = "MatClust")
    ```

2.  Use `vcov` to estimate the covariance matrix of the parameter estimates.

    The variance (covariance matrix) is computed straightforwardly:

    ```{r}
    vcov(mfit)
    ```

3.  Compare with the covariance matrix obtained when fitting a homogeneous Poisson model.

    ```{r}
    vcov(ppm(redwood ~ 1))
    ```

    As can be seen, the variance of the intensity estimate is quite a bit larger
    in the Matern model. This comes naturally by the doubly stochastic
    construction of the Matern model.

### Exercise 5

In this question we fit a Strauss point process model to
the `swedishpines` data.

1.  We need a guess at the interaction distance $R$. Compute and
    plot the $K$-function of the dataset and choose the value $r$
    which maximises the discrepancy $|K(r)-\pi r^2|$.

    We plot the above function which we want to maximize.

    ```{r}
    plot(Kest(swedishpines), abs(iso - theo) ~ r, main = "")
    ```

    As seen from the plot, the maximum lies around $r = 9$ by eye.
    We find the optimum explicitly like follows:

    ```{r}
    discrep <- function(r) {
      return(abs(as.function(Kest(swedishpines))(r) - pi*r^2))
    }
    res <- optimise(discrep, interval = c(0.1, 20), maximum = TRUE)
    print(res)
    R <- res$maximum
    ```
    
    This corresponds nicely with the plot.
    
2.  Fit the stationary Strauss model with the chosen interaction
    distance using
    ```{r, eval = FALSE}
    ppm(swedishpines ~ 1, Strauss(R))
    ```
    where `R` is your chosen value.
    
    As we have assigned `R`, we simply write:
    
    ```{r}
    fit <- ppm(swedishpines ~ 1, Strauss(R))
    ```

3.  Interpret the printout: how strong is the interaction?
    ```{r}
    print(fit)
    ```

    As seen, the $\gamma = `round(exp(coef(fit)[2]), 2)$ parameter is quite small. Thus there seems to be a
    strong negative association between points within distance R of each other.
    A $\gamma$ of $0$ implies the hard core process whereas $\gamma = 1$ implies
    the Poisson process and thus CSR.

4.  Plot the fitted pairwise interaction function using
    `plot(fitin(fit))`.

    The pairwise interaction function become:
    
    ```{r}
    plot(fitin(fit))
    ```

### Exercise 6

In Question 5 we guesstimated the Strauss interaction distance 
parameter. Alternatively this parameter could be estimated by 
profile pseudolikelihood.

1.  Look again at the plot of the $K$-function of
    `swedishpines` and determine a plausible range of
    possible values for the interaction distance.
    
    ```{r}
    plot(Kest(swedishpines), main = "")
    ```

    A conservative range of plausible interaction distances seems to be 5 to 12 meters.

2.  Generate a sequence of values equally spaced across this range,
    for example, if your range of plausible values was
    $[0.05, 0.3]$, then type
    ```{r, eval = FALSE}
    rvals <- seq(0.05, 0.3, by=0.01)
    ```

    We generate the numbers between 5 and 12.
    ```{r}
    rvals <- seq(5, 12, by = 0.1)
    ```

3.  Construct a data frame, with one column named `r`
    (matching the argument name of `Strauss`), containing
    these values. For example
    ```{r, eval = FALSE}
    D <- data.frame(r = rvals)
    ```

    OK, 
    
    ```{r}
    D <- data.frame(r = rvals)
    ```


4.  Execute
    ```{r, eval = FALSE}
    fitp <- profilepl(D, Strauss, swedishpines ~ 1)
    ```
    to find the maximum profile pseudolikelihood fit.

    OK, let's execute it: 
    
    ```{r}
    fitp <- profilepl(D, Strauss, swedishpines ~ 1)
    ```

5.  Print and plot the object `fitp`.

    ```{r}
    print(fitp)
    plot(fitp)
    ```

6.  Compare the computed estimate of interaction distance $r$ with
    your guesstimate. Compare the corresponding estimates of the
    Strauss interaction parameter $\gamma$.
    
    ```{r}
    (Ropt <- reach(as.ppm(fitp)))
    ```

    The $r = `r Ropt`$ is consistent with the previous guesstimate.

7.  Extract the fitted Gibbs point process model from the object
    `fitp` as 
    ```{r, eval = FALSE}
    bestfit <- as.ppm(fitp)
    ```

    OK, let's do that:

    ```{r}
    bestfit <- as.ppm(fitp)
    ```

### Exercise 7

Modify Question 5 by using the Huang-Ogata approximate
maximum likelihood algorithm (`method="ho"`) instead of maximum
pseudolikelihood (the default, `method="mpl"`).

```{r}
fit.mpl <- ppm(swedishpines ~ 1, Strauss(R), method = "mpl")
fit.ho  <- ppm(swedishpines ~ 1, Strauss(R), method = "ho")
print(fit.ho)
print(fit.mpl)
```

The fits are very similar.

### Exercise 8

Repeat Question 6 for the inhomogeneous Strauss process with
log-quadratic trend. The corresponding call to `profilepl` is
```{r}
fitp <- profilepl(D, Strauss, swedishpines ~ polynom(x,y,2))
```

```{r}
fitp2 <- profilepl(D, Strauss, swedishpines ~ polynom(x,y,2))
print(fitp2)
(bestfit <- as.ppm(fitp2))
reach(bestfit)
```
