---
title: 'Lab 2: Intensity -- solutions'
---

```{r, include = FALSE}
# This option turns evaluation of R code off in this document. Remove it
# if you use it as a template for solving the lab exercise.
knitr::opts_chunk$set(eval = TRUE)
```

This session covers exploratory tools
and formal model-fitting procedures for investigating intensity.

```{r, message=FALSE}
library(spatstat)
```

### Exercise 1

The dataset `japanesepines` contains the locations of
Japanese Black Pine trees in a study region.

1.  Plot the `japanesepines` data.

    We use the generic `plot` function which is dispatched to `plot.ppp`:
    
    ```{r}
    plot(japanesepines)
    ```

2.  What is the average intensity (the average number of points
    per unit area?

    The average intensity can be computed via `intensity.ppp`:
    
    ```{r}
    intensity(japanesepines)
    ```

3.  Using `density.ppp`, compute a kernel estimate of the
    spatially-varying intensity function for the Japanese pines
    data, using a Gaussian kernel with standard deviation
    $\sigma=0.1$ units, and store the estimated intensity in an
    object `D` say.

    From the documentation (`?density.ppp`) we see that the following will work:
    
    ```{r}
    D <- density(japanesepines, sigma = 0.1)
    ```

4.  Plot a colour image of the kernel estimate `D`.

    The plotting of the colour image is automatically done by dispatched call to
    the `plot.im` method by calling `plot` on the `im` object.

    ```{r}
    plot(D, main = "")
    ```

5.  Most plotting commands will accept the argument
    `add=TRUE` and interpret it to mean that the plot
    should be drawn over the existing display, without clearing the
    screen beforehand. Use this to plot
    a colour image of the kernel estimate `D` with
    the original Japanese Pines data superimposed.

    We can use the `add = TRUE` functionality of the plotting methods.
    
    ```{r}
    plot(D, main = "")
    plot(japanesepines, add = TRUE, cols = "white", cex = 0.5, pch = 16)
    ```

6.  Plot the kernel estimate without the ‘colour ribbon’.

    From `help("plot.im")` we see that `ribbon = FALSE` disables the colour key:
    ```{r}
    plot(D, main = "", ribbon = FALSE)
    plot(japanesepines, add = TRUE, cols = "white", cex = 0.5, pch = 16)
    ```

7.  Try the following command
    ```{r}
    persp(D, theta=70, phi=25, shade=0.4)
    ```
    and find the documentation for the arguments `theta`,
    `phi` and `shade`.

    It dispatches to `persp.im`, but these arguments are then passed down to
    `persp.default` through the dots (`...`).
    From the documentation of `persp.default` they are "angles defining the
    viewing direction. `theta` gives the azimuthal direction and `phi` the
    colatitude."
    The `shade` controls the shading of the surface facets.

    ```{r}
    persp(D, theta=70, phi=25, shade=0.4, main = "")
    ```

7.  Find the maximum and minimum values of the intensity estimate
    over the study region. (Hint: Use `summary` or `range`)
    
    ```{r}
    range(D)
    ```

8.  The kernel estimate of intensity is defined so that its integral
    over the entire study region is equal to the number of points in
    the data pattern, ignoring edge effects. Check whether this is
    approximately true in this example. (Hint: use `integral`)
    
    Calling `integral.im` we see that the integral is close to the
    observed number of points `r npoints(japanesepines)`:
    
    ```{r}
    round(integral(D))
    ```

### Exercise 2

The `bei` dataset gives the locations of trees in a survey area with additional
covariate information in a list `bei.extra`.

1.  Assign the elevation covariate to a variable `elev` by typing
    ```{r}
    elev <- bei.extra$elev
    ```

2.  Plot the trees on top of an image of the elevation covariate.
    ```{r}
    plot(elev)
    plot(bei, add = TRUE, col = "black")
    ```

3.  Assume that the intensity of trees is a function $\lambda(u) = \rho(e(u))$
    where $e(u)$ is the terrain elevation at location u.
    Compute a nonparametric estimate of the function $\rho$ and plot it by
    ```{r}
    rh <- rhohat(bei, elev)
    plot(rh)
    ```

4.  Compute the predicted intensity based on this estimate of $\rho$.
    
    ```{r}
    prh <- predict(rh)
    plot(prh, main = "")
    plot(bei, add = TRUE, cols = "white", cex = .2, pch = 16)
    ```

5.  Compute a non-parametric estimate of intensity by kernel smoothing, and
  	compare with the predicted intensity above.

    The kernel density estimate of the points is computed and plotted with the
    following code:
    
    ```{r}
    dbei <- density(bei, sigma = bw.scott)
    plot(dbei, main = "")
    plot(bei, add = TRUE, cols = "white", cex = .2, pch = 16)
    ```
    
    Which seems to be quite different form the predicted intentisty.

6.  Bonus info: To plot the two intensity estimates next to each
  	other you collect the estimates as a spatial object list
  	(`solist`) and plot the result (the estimates are called
  	`pred` and `ker` below):
    ```{r eval=FALSE}
    l <- solist(pred, ker)
    plot(l, equal.ribbon = TRUE, main = "", 
         main.panel = c("rhohat prediction", "kernel smoothing"))
    ```

    ```{r}
    l <- solist(prh, dbei)
    plot(l, equal.ribbon = TRUE, main = "",
         main.panel = c("rhohat prediction", "kernel smoothing"))
    ```


### Exercise 3

The command `rpoispp(100)` generates realisations of the
Poisson process with intensity $\lambda = 100$ in the unit square.

1.  Repeat the command `plot(rpoispp(100))` several times
    to build your intuition about the appearance of a completely
    random pattern of points.

    Let's plot it three times:
    
    ```{r, results='hide'}
    replicate(3, plot(rpoispp(lambda = 100), main = ""))
    ``` 

    As can be seen, the points (unsurprisingly) are much more random that want one 
    might think. "Randomly" drawing points on a piece of paper one would usually
    draw a point pattern that is more regular (i.e. the points are repulsive).

2.  Try the same thing with intensity $\lambda = 1.5$.

    For brevity we only do it once here:
    
    ```{r, results='hide'}
    plot(rpoispp(lambda = 1.5), main = "")
    ``` 

    Here we expect 1.5 points in the plot each time.


### Exercise 4

Returning to the Japanese Pines data,

1.  Fit the uniform Poisson point process model to the Japanese
    Pines data
    ```{r, eval = FALSE}
    ppm(japanesepines~1)
    ```

    We fit the Poisson process model with the given command and print the output:
    
    ```{r}
    m.jp <- ppm(japanesepines ~ 1)
    print(m.jp)
    ```

2.  Read off the fitted intensity. Check that this is the correct
    value of the maximum likelihood estimate of the intensity.

    We extract the coeficient with the `coef` function, and compare to the
    straightforward estimate obtained by `intensity``:

    ```{r}
    unname(exp(coef(m.jp)))
    intensity(japanesepines)
    ```

    As seen, they agree exactly.

### Exercise 5

The `japanesepines` dataset is believed to exhibit
spatial inhomogeneity.

1.  Plot a kernel smoothed intensity estimate.

    Plot the kernel smoothed intensity estimate selecting the bandwidth with `bw.scott`:
    
    ```{r}
    jp.dens <- density(japanesepines, sigma = bw.scott)
    plot(jp.dens)
    plot(japanesepines, col = "white", cex = .4, pch = 16, add = TRUE)
    ```

2.  Fit the Poisson point process models with loglinear intensity
    (trend formula `~x+y`) and log-quadratic intensity (trend
    formula `~polynom(x,y,2)`) to the Japanese Pines data.

    We fit the two models with `ppm`:
    
    ```{r}
    jp.m <- ppm(japanesepines ~ x + y)
    jp.m2 <- ppm(japanesepines ~ polynom(x, y, 2) )
    ```

3.  extract the fitted coefficients for these models using
    `coef`.
        
    ```{r}
    coef(jp.m)
    coef(jp.m2)
    ```
    
4.  Plot the fitted model intensity (using `plot(predict(fit))`)

    ```{r}
    par(mar=rep(0,4))
    plot(predict(jp.m), main = "")
    plot(predict(jp.m, se=TRUE)$se, main = "")
    
    plot(predict(jp.m2), main = "")
    plot(predict(jp.m2, se=TRUE)$se, main = "")
    ```

5.  perform the Likelihood Ratio Test for the null hypothesis of a
    loglinear intensity against the alternative of a log-quadratic
    intensity, using `anova`.

    ```{r}
    anova(jp.m, jp.m2)
    ```

6.  Generate 10 simulated realisations of the fitted log-quadratic
    model, and plot them, using `plot(simulate(fit, nsim=10))` where 
    `fit` is the fitted model.

    ```{r}
    par(mar=rep(0.5,4))
    plot(simulate(jp.m2, nsim=10), main = "")
    ```

### Exercise 6

The `update` command can be used to re-fit a point
process model using a different model formula.

1.  Type the following commands and interpret the results:
    ```{r}
    fit0 <- ppm(japanesepines ~ 1)
    fit1 <- update(fit0, . ~ x)
    fit1
    fit2 <- update(fit1, . ~ . + y)
    fit2
    ```

    OK, let's do that:
    
    ```{r}
    fit0 <- ppm(japanesepines ~ 1)
    fit1 <- update(fit0, . ~ x)
    fit1
    fit2 <- update(fit1, . ~ . + y)
    fit2
    ```

2.  Now type `step(fit2)` and interpret the results.

    The backwards selection is done with the code:
    
    ```{r}
    step(fit2)
    ```
    
    First, given two models the preferred model is the one with the minimum AIC 
    value. In step 1, the removal of x results in the least AIC and is hence 
    deleted. In step 2, removing y results in a lower AIC than not deleing anything 
    and is thus deleted. This results in the constant model.
    
### Exercise 7

The `bei` dataset gives the locations of trees in a survey area
with additional covariate information in a list `bei.extra`.

1.  Fit a Poisson point process model to the data which assumes
    that the intensity is a loglinear function of terrain slope
    and elevation (hint: use `data = bei.extra` in `ppm`).

    We fit the log-linear intensity model with the following:
    
    ```{r}
    bei.m <- ppm(bei ~ elev + grad, data = bei.extra)
    ```

2.  Read off the fitted coefficients and write down the fitted
    intensity function.

    The coefficents are extraced with `coef`:
    
    ```{r}
    coef(bei.m)
    ```

    Hence the model is $log\lambda(u) = -8.55 + 0.02\cdot E(u) + 5.84 G(u)$ where
    $E(u)$ and $G(u)$ is the elevation and gradient, respectively, at $u$.

3.  Plot the fitted intensity as a colour image.

    ```{r}
    plot(predict(bei.m), main = "")
    plot(bei, cex = 0.3, pch = 16, cols = "white", add = TRUE)
    ```

4.  extract the estimated variance-covariance matrix of the
    coefficient estimates, using `vcov`.

    We call `vcov` on the fitted model object:
    ```{r}
    vcov(bei.m)
    ```

5.  Compute and plot the standard error of the intensity estimate
    (see `help(predict.ppm)`).

    From the documentation the argument `se` will trigger the computation of the 
    standard errors. These are then plotted in the standard manner.
    
    ```{r}
    std.err <- predict(bei.m, se = TRUE)$se
    plot(std.err, main = "")
    ```

### Exercise 8

Fit Poisson point process models to the Japanese Pines data, with
the following trend formulas. Read off an expression for the fitted
intensity function in each case.

|Trend formula     |Fitted intensity function                   |
|:-----------------|:-------------------------------------------|
|`~1`              | $\log\lambda(u) = 4.17$                    |
|`~x`              | $\log\lambda(u) = 4.28 - 0.23x$            |
|`~sin(x)`         | $\log\lambda(u) = 4.29 - 0.26\sin(x)$      |
|`~x+y`            | $\log\lambda(u) = 4.07 - 0.23x + 0.42y$    | 
|`~polynom(x,y,2)` | $\log\lambda(u) = 4.06 + 1.14x - 1.56y - 0.75x^2 - 1.20xy + 2.51y^2$ |
|`~factor(x < 0.4)`| $\log\lambda(u) = 4.10 + 0.16\cdot I(x < 0.4)$ |

(Here, $I(\cdot)$ denote the indicator function.)

The fitted intensity functions have been written into the table based on the follwing model fits:

```{r}
coef(ppm1 <- ppm(japanesepines ~ 1)) 
coef(ppm2 <- ppm(japanesepines ~ x))
coef(ppm3 <- ppm(japanesepines ~ sin(x)))
coef(ppm4 <- ppm(japanesepines ~ x + y))
coef(ppm5 <- ppm(japanesepines ~ polynom(x, y, 2)))
coef(ppm6 <- ppm(japanesepines ~ factor(x < 0.4)))
```

### Exercise 9

Make image plots of the fitted intensities for the inhomogeneous
models above.

Again, we use `plot(predict())`:

```{r}
plot(predict(ppm1), main = "")
plot(predict(ppm2), main = "")
plot(predict(ppm3), main = "")
plot(predict(ppm4), main = "")
plot(predict(ppm5), main = "")
plot(predict(ppm6), main = "")
```

### Exercise 10

The dataset `hamster` is a multitype pattern representing
the locations of cells of two types, *dividing* and *pyknotic*.

1.  plot the patterns of pyknotic and dividing cells separately;

    ```{r}
    plot(split(hamster), main = "")
    ```

2.  plot kernel estimates of the intensity functions of pyknotic and
    dividing cells separately;

    ```{r}
    plot(density(split(hamster)), main = "")
    ```


3.  use `relrisk` to perform cross-validated bandwidth
    selection and computation of the relative intensity of pyknotic
    cells.

    ```{r}
    plot(relrisk(hamster, hmax = 1, relative = TRUE, control = "dividing"))
    ```

### Exercise 11

The dataset `ants` is a multitype point pattern representing the
locations of nests of two species of ants.

1.  plot the data.

    ```{r}
    plot(ants)
    ```

2.  Fit the model `ppm(ants ~ marks)` and interpret the result.
    Compare the result with `summary(ants)` and explain
    the similarities.

    ```{r}
    fit1 <- ppm(ants ~ marks)
    ```
    
    This is a Poisson model with a separate constant intensity for each mark. The
    fitted intensities are:
    ```{r}
    exp(coef(fit1)[1])
    exp(coef(fit1)[1] + coef(fit1)[2])
    ```

    This agrees perfectly with the output of `summary(ants)`:
    ```{r}
    summary(ants)
    ```


3.  Fit the model `ppm(ants ~ marks + x)` and write down an
    expression for the fitted intensity function.

    ```{r}
    fit2 <- ppm(ants ~ marks + x)
    (co <- coef(fit2))
    ```
    
    Intensity for the reference type (Cataglyphis):
    
    $$\lambda( (x,y) ) = \exp(`r co[["(Intercept)"]]` + `r co[["x"]]` \cdot x)$$

    Intensity for the other type (Messor):

    $$\lambda( (x,y) ) = \exp(`r co[["(Intercept)"]]` + `r co[["marksMessor"]]` + `r co[["x"]]` \cdot x)$$


4.  Fit the model `ppm(ants ~ marks * x)` and write down an
    expression for the fitted intensity function.

    ```{r}
    fit3 <- ppm(ants ~ marks * x)
    (co <- coef(fit3))
    ```
    
    Intensity for the reference type (Cataglyphis):
    
    $$\lambda( (x,y) ) = \exp(`r co[["(Intercept)"]]` + `r co[["x"]]` \cdot x)$$
    
    Intensity for the other type (Messor):

    $$\lambda( (x,y) ) = \exp(`r co[["(Intercept)"]]` + `r co[["marksMessor"]]` + (`r co[["x"]]` + `r co[["marksMessor"]]`) \cdot x)$$

5.  Compute the fitted intensities of the three models fitted above 
    using `predict` and plot the results.

    ```{r}
    pred <- c(predict(fit1), predict(fit2), predict(fit3))
    plot(as.solist(pred), ncols = 2, main = "")
    ```

6.  Explain the difference between the models fitted by
    `ppm(ants ~ marks + x)` and `ppm(ants ~ marks * x)` .

    For the additive model the effect of the `x` coordinate is the same for both types
    of ants, while the effect of `x` differs in the multiplicative model.

## Exercise 12

The study region for the ants’ nests data `ants` is
divided into areas of "scrub" and "field". We want to fit a Poisson
model with different intensities in the field and scrub areas.

The coordinates of two points on the boundary line between field and
scrub are given in `ants.extra$fieldscrub`. First construct a function that
determines which side of the line we are on:

```{r}
fs <- function(x,y) {
  ends <- ants.extra$fieldscrub
  angle <- atan(diff(ends$y)/diff(ends$x))
  normal <- angle + pi/2
  project <- (x - ends$x[1]) * cos(normal) + (y - ends$y[1]) * sin(normal)
  factor(ifelse(project > 0, "scrub", "field"))
}
```

Now fit the models:
```{r, eval=FALSE}
ppm(ants ~ marks + side, data = list(side=fs))
ppm(ants ~ marks * side, data = list(side=fs))
```
and interpret the results.

```{r}
fit1 <- ppm(ants ~ marks + side, data = list(side=fs))
fit2 <- ppm(ants ~ marks * side, data = list(side=fs))
```

In the first model the fitted intensity is lower in the scrub than in the field
(but this effect is not significant).

In the second model the fitted intensity of Cataglyphis is lower in the scrub
than the intensity of Cataglyphis in the field, where as it is the other way
around for Messor. When we allow for the different effect between ant types the
scrub/field covariate is significant.