---
title: 'OGH talk: Analysing spatial point patterns with spatstat'
author: "Ege Rubak"
date: "2 September 2021"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Plan for the session

- Very short introduction (these slides).

- Live tutorial with examples of analysis.

- Questions and discussion.

## The spatstat package

-   25 years development by mainly Adrian Baddeley with co-author Rolf
    Turner from the beginning and more recently I joined the team.
-   100,000+ lines of R code, 30,000+ lines of C code, 2,500+ exported
    objects and 1000+ documentation files.
-   Recently split into sub-packages `spatstat.xxxx`, which the now
    almost empty package `spatstat` then `Depends` on.

## The spatstat design

-   Strong focus on backwards compatibility.
-   Follows base R design and uses base R graphics.
-   Extensive usage of the S3 object system.
-   Has its own S3 objects for data etc.
-   Detailed documentation.
-   Use `?spatstat` to get an overview and find hidden gems.

## What to use spatstat for?

-   You **can use** `spatstat` to describe/summarise any given point set with
    things like

    -   Pairwise distances, nearest neighbour distances, empty space
        distances, Dirichlet/Voronoi tessellations, ...

-   However, `spatstat` really focuses on **statistical inference** for
    phenomena that generate **random locations** (point processes).
    
## Complete spatial randomness (Poisson process)

```{r message=FALSE, fig.width=10, fig.height=4}
library(spatstat)
set.seed(42) # Reproducibility
Xpois <- rpoispp(100, nsim = 3)
plot(Xpois, main = "")
```

## Hard core Gibbs process

```{r message=FALSE, fig.width=10, fig.height=4}
Xhc <- rHardcore(beta = 100, R = .05, nsim = 3)
plot(Xhc, main = "")
```

## Thomars cluster process

```{r message=FALSE, fig.width=10, fig.height=4}
Xthomas <- rThomas(kappa = 5, mu = 20, scale = .1, nsim = 3)
plot(Xthomas, main = "")
```

## Inhomogeneous Poisson process

```{r message=FALSE, fig.width=10, fig.height=4}
lambda <- function(x,y){200*(x^2+y^2)}
Xinhom <- rpoispp(lambda, nsim = 3)
plot(Xinhom, main = "")
```

## Separating intensity and interaction

- Intensity is a first moment property.

- Interaction is a higher moment property (inter-point correlation).

- They are confounded and without further assumptions it is impossible to
  separate them in general.
  
- Often a approach like in time series is used:
    + First, model the mean (trend, seasonality).
    + Second, model the interaction after accounting for the mean model.
    
## How to learn more and report bugs

-   Explore `?spatstat` which includes lists of commonly (and less
    commonly) used functions.

-   Get the book. Unfortunately we don't have a license to share an
    online version as many authors have nowadays. Maybe this will change
    with a second edition. There are three free sample chapters at
    <https://book.spatstat.org/>

-   Ask questions on stackoverflow under the [spatstat
    tag](https://stackoverflow.com/tags/spatstat/)

-   Report bugs or make feature requests on
    [GitHub](https://github.com/spatstat/spatstat). (If possible find the
    right sub package repo to put the issue under.)
