---
title: 'OGH tutorial: Analysing spatial point patterns with spatstat'
author: "Ege Rubak"
date: "2 September 2021"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float:  
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This tutorial is for the 2021 OpenGeoHub summer school. It is written with very 
brief explanations of the content. Please watch the recorded tutorial 
presentation for detailed explanations of the content.

To follow along you need `spatstat` installed along with its dependencies.
You can install the development version like this (but the CRAN version is fine for this tutorial):
```{r eval=FALSE}
options(repos = c(
    spatstat = 'https://spatstat.r-universe.dev',
    CRAN = 'https://cloud.r-project.org'))
install.packages("spatstat", dependencies = TRUE)
```

Now load spatstat and the needed sub-packages:
```{r}
library(spatstat)
```

## Point pattern data

Planar coordinates (x,y) are assumed almost everywhere in
`spatstat`, so **geographic (lng,lat) coordinates must be projected**. You can
use [`sf`](https://github.com/r-spatial/sf) and <https://projectionwizard.org> 
for this (in rare cases `sp` and `maptools` may be useful).

As a first example consider the built-in Chorley-Ribble cancer data:

```{r}
plot(chorley, cols = c("red", "blue"))
```

It is stored as planar point pattern `ppp` with many S3 methods available
for plotting, summarising, subsetting, etc.

```{r}
class(chorley)
methods(class = "ppp")
summary(chorley)
```

**The spatial region where the points were observed (observation
window, `owin`) is part of the `ppp` object.**
There are many polygonal operations available to manipulate these (handled by `polyclip` package). E.g:

```{r fig.width=12, fig.height=5}
W <- Window(chorley)
Wplus <- dilation(W, 2)
Wminus <- erosion(W, 2)
plot(Wplus, main = "Dilation and erosion")
plot(W, add = TRUE)
plot(Wminus, add = TRUE)
```

# Case study: Murchison gold deposits

-   Objective: Determine gold prospectivity from predictors which are
    easier to observe than gold deposits directly.

```{r}
murchison
plot(murchison, equal.scales = TRUE, main = "")
```

Rescaling to km:

```{r}
gold <- rescale(murchison$gold, 1000, "km")
faults <- rescale(murchison$faults, 1000, "km")
gs <- rescale(murchison$greenstone, 1000, "km")
```

Distance to fault lines (resulting raster image is called an `im` in
`spatstat`):

```{r}
dfaults <- distmap(faults, eps = 1) # 1km resolution
plot(dfaults)
plot(faults, add = TRUE, col = "white")
plot(gold, add = TRUE, cols = "gold")
```

## Relationship between gold intensity and fault line distance

Dividing space according to distance (tessellation according to
quantiles):

```{r}
qt <- quantess(Window(gold), dfaults, 4)
plot(qt, las = 1) # horizontal labels on axis (colourmap)
```

Deposits in each band around fault lines:

```{r}
gold4 <- quadratcount(gold, tess = qt)
gold4
barplot(intensity(gold4))
```

```{r}
gold8 <- quadratcount(gold, tess = quantess(Window(gold), dfaults, 8))
gold8
barplot(intensity(gold8))
```

Kernel smoothed version (estimate of true functional relationship
between covariate and intensity with pointwise CI):

```{r}
rh <- rhohat(gold, dfaults)
plot(rh, legendpos = "topright", xlim = c(0,30))
```

Predictions from this relationship compared to kernel smoothed
intensity:

```{r}
pred_list <- solist(
  rhohat = predict(rh),
  kernel = density(gold, sigma = 5),
  adaptive = adaptive.density(gold, method = "kernel")
)
plot(pred_list, equal.ribbon = TRUE)
```

## ROC curve

Another view is to compare fault distances for observed gold deposits
with fault distances for uniformly random points.

First we need to extract values of the distmap at gold locations (we can use `[`
to extract a spatial subset):

```{r}
dgold <- dfaults[gold]
head(dgold)
```

More precise values are obtained by using `distfun()`:

```{r}
dfaults_fun <- distfun(faults)
dfaults_fun
dgold <- dfaults_fun(gold)
head(dgold)
```

Comparing histograms:

```{r}
rand <- runifpoint(ex = gold)
drand <- dfaults_fun(rand)
hist(drand, col = rgb(.8,0,0,.1), ylim = c(0,150))
hist_dgold <- hist(dgold, plot = FALSE)
plot(hist_dgold, col = rgb(0,0,.8,.1), add = TRUE)
```

In terms of the ROC:

```{r}
plot(roc(gold, dfaults_fun, high = FALSE))
```

Interpretation: If you explore the 10% of the region that is closest to
fault lines you find \~60% of the gold deposits. At 20% it is \~80%.

## Poisson models

**Log**-linear models for the intensity follow the syntax of `lm`/`glm`:

```{r}
(model_dist <- ppm(gold ~ dfaults_fun, eps=1))
```

The fitted model is of the form
$$
\lambda(u) = \exp( \alpha + \beta d(u) ).  
$$
That is, the estimated intensity of gold deposits in the immediate
vicinity of a geological fault is about
$\exp(\hat{\alpha}) = `r exp(coef(model_dist)[1])`$ and decreases by a
*factor* of $\exp(\hat{\beta}) = `r exp(coef(model_dist)[2])`$ for every
additional kilometre away from a fault.

Use `effectfun()` to see the effect of the distance covariate on the
intensity:

```{r}
ef_dist <- effectfun(model_dist, "dfaults_fun", se.fit = TRUE)
plot(ef_dist)
```

Use `simulate()` to generate realizations from the model:

```{r}
s_dist <- simulate(model_dist, nsim = 8)
s_dist[[9]] <- gold
names(s_dist)[9] <- "gold"
plot(s_dist)
```

Models with two covariates:

```{r}
(model_add <- ppm(gold ~ dfaults_fun + gs, eps=1))
(model_int <- ppm(gold ~ dfaults_fun * gs, eps=1))
```

Formal test between nested models:

```{r}
anova(model_add, model_int, test = "Chi")
```

Simulations from interaction model:

```{r}
s_int <- simulate(model_int, nsim = 8)
s_int[[9]] <- gold
names(s_int)[9] <- "gold"
plot(s_int)
```

## Model diagnostics

The classical tools from GLMs have extensions to point process models:

```{r}
lev <- leverage(model_int)
inf <- influence(model_int)
dfb <- dfbetas(model_int)
```

High leverage of areas within greenstone far from faults. I.e. adding and 
deleting points in this area has a large effect on the fitted model.
```{r}
plot(lev, showcut = FALSE)
plot(faults, add = TRUE, col = "white")
plot(gs, add = TRUE, border = "green", )
```

Most influential data points are in greenstone far from fault lines:
```{r}
plot(inf)
plot(faults, add = TRUE, col = "grey")
plot(gs, add = TRUE, border = "green", )
```

The sign of the effect on each parameter can be found by studying `dfbetas()`:
```{r}
plot(dfb)
```

# Other models and simulation

## Log-Gaussian Cox Process

This uses the `RandomFields` package internally, so it must be
installed.

```{r}
model_clust <- kppm(gold ~ dfaults*gs, clusters = "LGCP", eps = 2, method = "clik2")
```

Notice how the standard errors now are larger than the Poisson case:

```{r}
coef(summary(model_clust))
coef(summary(model_int))
```

Simulations from the model:

```{r}
s_clust <- simulate(model_clust, nsim = 8)
s_clust[[9]] <- gold
names(s_clust)[9] <- "gold"
plot(s_clust)
```

## Gibbs models for patterns with inhibition

- Synthetic example:

```{r}
X <- rStraussHard(beta = 200, gamma = 0.4, R = 0.06, H = 0.03, nsim = 9)
plot(X)
```

- Model fitting using so-called pseudo-likelihood:

```{r}
(model_gibbs <- ppm(X[[1]] ~ 1, interaction = StraussHard(r = 0.06, hc = 0.03)))
```

Ripley's K-function is the normalized expected number of further points within distance r given that you are at a point of the process:

```{r}
plot(Kest(X[[1]]))
```

The pair-correlation function is the derivative and has the advantage of not being cumulative:

```{r}
plot(pcf(X[[1]]))
```


# Appendix: Other popular capabilities of spatstat

## Simulation envelopes

Default simulation under homogeneous Poisson:
```{r}
env_gibbs_wrong <- envelope(X[[1]], Kest)
plot(env_gibbs_wrong, .-pi*r^2 ~ r)
```

Simulation from the model:
```{r}
env_gibbs_right <- envelope(model_gibbs, Kest)
plot(env_gibbs_right, .-pi*r^2 ~ r)
```

## Linear networks

Street crimes in Chicago:
```{r}
chicago
plot(chicago)
```

Intensity of crimes:
```{r}
lam <- density(chicago, sigma = 100)
plot(lam)
```

```{r}
lam2 <- densityVoronoi(chicago, nrep = 50, f = .2)
plot(lam2)
```

See the session by Andrea Gilardi for much more on networks.

## Tessellations

Dirichlet tesselation of gold pattern:
```{r}
gold_tess <- dirichlet(gold)
plot(gold_tess)
```

Dirichlet tessellation on network for Chicago crimes:
```{r}
XX <- chicago
Window(XX) <- grow.rectangle(Window(chicago), 1) ## Needed due to a minor bug
chicago_tess <- lineardirichlet(XX)
plot(chicago_tess)
```
