---
title: 'Lab 2: Intensity'
---

```{r, include = FALSE}
# This option turns evaluation of R code off in this document. Remove it
# if you use it as a template for solving the lab exercise.
knitr::opts_chunk$set(eval = FALSE)
```

This session covers exploratory tools for investigating intensity.  

```{r, include=FALSE}
library(spatstat)
```

### Exercise 1

The dataset `japanesepines` contains the locations of
Japanese Black Pine trees in a study region.

1.  Plot the `japanesepines` data.

2.  What is the average intensity (the average number of points
    per unit area?

3.  Using `density.ppp`, compute a kernel estimate of the
    spatially-varying intensity function for the Japanese pines
    data, using a Gaussian kernel with standard deviation
    $\sigma=0.1$ units, and store the estimated intensity in an
    object `D` say.

4.  Plot a colour image of the kernel estimate `D`.

5.  Most plotting commands will accept the argument
    `add=TRUE` and interpret it to mean that the plot
    should be drawn over the existing display, without clearing the
    screen beforehand. Use this to plot
    a colour image of the kernel estimate `D` with
    the original Japanese Pines data superimposed.

6.  Plot the kernel estimate without the ‘colour ribbon’.

7.  Try the following command
    ```{r}
    persp(D, theta=70, phi=25, shade=0.4)
    ```
    and find the documentation for the arguments `theta`,
    `phi` and `shade`.

7.  Find the maximum and minimum values of the intensity estimate
    over the study region. (Hint: Use `summary` or `range`)

8.  The kernel estimate of intensity is defined so that its integral
    over the entire study region is equal to the number of points in
    the data pattern, ignoring edge effects. Check whether this is
    approximately true in this example. (Hint: use `integral`)

### Exercise 2

The `bei` dataset gives the locations of trees in a survey area with additional
covariate information in a list `bei.extra`.

1.  Assign the elevation covariate to a variable `elev` by typing
    ```{r}
    elev <- bei.extra$elev
    ```

2.  Plot the trees on top of an image of the elevation covariate.

3.  Assume that the intensity of trees is a function $\lambda(u) = \rho(e(u))$
    where $e(u)$ is the terrain elevation at location u.
    Compute a nonparametric estimate of the function $\rho$ and plot it by
    ```{r}
    rh <- rhohat(bei, elev)
    plot(rh)
    ```

4.  Compute the predicted intensity based on this estimate of $\rho$.

5.  Compute a non-parametric estimate of intensity by kernel smoothing, and
  	compare with the predicted intensity above.

6.  Bonus info: To plot the two intensity estimates next to each
  	other you collect the estimates as a spatial object list
  	(`solist`) and plot the result (the estimates are called
  	`pred` and `ker` below):
    ```{r}
    l <- solist(pred, ker)
    plot(l, equal.ribbon = TRUE, main = "", 
         main.panel = c("rhohat prediction", "kernel smoothing"))
    ```

### Exercise 3

The command `rpoispp(100)` generates realisations of the
Poisson process with intensity $\lambda = 100$ in the unit square.

1.  Repeat the command `plot(rpoispp(100))` several times
    to build your intuition about the appearance of a completely
    random pattern of points.

2.  Try the same thing with intensity $\lambda = 1.5$.


### Exercise 4

Returning to the Japanese Pines data,

1.  Fit the uniform Poisson point process model to the Japanese
    Pines data
    ```{r}
    ppm(japanesepines~1)
    ```

2.  Read off the fitted intensity. Check that this is the correct
    value of the maximum likelihood estimate of the intensity.

### Exercise 5

The `japanesepines` dataset is believed to exhibit
spatial inhomogeneity.

1.  Plot a kernel smoothed intensity estimate.

2.  Fit the Poisson point process models with loglinear intensity
    (trend formula `~x+y`) and log-quadratic intensity (trend
    formula `~polynom(x,y,2)`) to the Japanese Pines data.

3.  extract the fitted coefficients for these models using
    `coef`.

4.  Plot the fitted model intensity (using `plot(fit)`)

5.  perform the Likelihood Ratio Test for the null hypothesis of a
    loglinear intensity against the alternative of a log-quadratic
    intensity, using `anova`.

6.  Generate 10 simulated realisations of the fitted log-quadratic
    model, and plot them, using `plot(simulate(fit, nsim=10))` where 
    `fit` is the fitted model.


### Exercise 6

The `update` command can be used to re-fit a point
process model using a different model formula.

1.  Type the following commands and interpret the results:
    ```{r}
    fit0 <- ppm(japanesepines ~ 1)
    fit1 <- update(fit0, . ~ x)
    fit1
    fit2 <- update(fit1, . ~ . + y)
    fit2
    ```

2.  Now type `step(fit2)` and interpret the results.
    

### Exercise 7

The `bei` dataset gives the locations of trees in a survey area
with additional covariate information in a list `bei.extra`.

1.  Fit a Poisson point process model to the data which assumes
    that the intensity is a loglinear function of terrain slope
    and elevation (hint: use `data = bei.extra` in `ppm`).

2.  Read off the fitted coefficients and write down the fitted
    intensity function.

3.  Plot the fitted intensity as a colour image.

4.  extract the estimated variance-covariance matrix of the
    coefficient estimates, using `vcov`.

5.  Compute and plot the standard error of the intensity estimate
    (see `help(predict.ppm)`).


### Exercise 8

Fit Poisson point process models to the Japanese Pines data, with
the following trend formulas. Read off an expression for the fitted
intensity function in each case.

|Trend formula     |Fitted intensity function                   |
|:-----------------|:-------------------------------------------|
|`~1`              |                                            |
|`~x`              |                                            |
|`~sin(x)`         |                                            |
|`~x+y`            |                                            |
|`~polynom(x,y,2)` |                                            |
|`~factor(x < 0.4)`|                                            |

### Exercise 9

Make image plots of the fitted intensities for the inhomogeneous
models above.
