---
title: "Notes for session 2"
author: "Adrian Baddeley and Ege Rubak"
date: "2019-07-02"
output: html
---

# Intensity


## Intensity and probability density

Definition

Objectives

## Nonparametric estimation

### Nonparametric estimation of spatially-varying intensity

```{r}
library(spatstat)
X <- japanesepines
A <- quadratcount(X, 3, 3)
Z <- density(X)
Z <- density(X, bw.diggle)
plot(Z, main="")
```

### Nonparametric estimation of spatially-varying, mark-dependent intensity

```{r}
M <- split(mucosa)
B <- density(M, bw.diggle)
plot(B, main="")
```

Ratio of intensities

```{r}
plot(relrisk(mucosa, casecontrol=FALSE))
```

### Nonparametric estimation of intensity depending on a covariate

```{r}
E <- split(mucosa)$ECL
g <- rhohat(E, "y")
plot(g)
```

```{r}
X <- murchison$gold
L <- murchison$faults
X <- rescale(X, 1000, "km")
L <- rescale(L, 1000, "km")
D <- distfun(L)
h <- rhohat(X, D)
plot(h)
```

## Parametric modelling

### Loglinear model for intensity

$$\log\lambda(u) = \beta_1 Z_1(u) + \ldots + \beta_p Z_p(u)$$

### Fit using Poisson likelihood

```{r}
fit <- ppm(X ~ D)
coef(fit)
plot(effectfun(fit, "D"), xlim=c(0, 20))
```
