---
title: "Notes for session 3"
author: "Adrian Baddeley and Ege Rubak"
date: "July 15, 2019"
---

# Dependence between points

Another important goal is to detect stochastic dependence between
points in a point pattern.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=5}
library(spatstat)
Xrandom <- rpoispp(50)
Xlist <- solist(inhibition=cells,
		random=Xrandom,
		clustering=redwood)
plot(Xlist, main="")		
```

* The homogeneous Poisson process is regarded as completely random
(Complete Spatial Randomness, CSR).
* A point process is called
_inhibited_ if the distances between points are typically _larger_ than
expected for a random pattern with the same intensity.
* A point process is called _clustered_
if the distances between points are typically _smaller_ than would be
expected for a random pattern with the same intensity.

The terms _inhibited_ and _clustered_ are analogous, respectively, to
"negatively correlated" and "positively correlated". They do not imply
any particular kind of stochastic dependence and they do not explain how
the pattern was generated. 

Dependence between points is sometimes called _"interaction"_, but this term
is dangerous because it suggests a particular mechanism for the
dependence. 

## Exploratory tools

Traditionally (1976-2005) dependence between points was investigated
using exploratory tools such as Ripley's $K$ function
and the pair correlation function. While these methods are useful
and convenient, they have limitations. They have become so popular that
many user communities are not interested in learning better methods
that have been developed in recent years.
In this workshop, as an experiment we are reducing the coverage of these
methods to a minimum, so that we can concentrate on the more modern methods.

### K-function

The (Ripley) $K$-function assumes the point process has
constant intensity $\lambda$. It is defined so that,
for a typical random point in the point process,
the number of other random points lying closer than a distance $r$
has expected value $\lambda \, K(r)$.

For a completely random (homogeneous
Poisson) process, $K(r) = \pi r^2$. An inhibited process will usually
have $K(r) < \pi r^2$, while a clustered process will have
$K(r) > \pi r^2$, for appropriate values of $r$.

An estimate of the $K$ function can be computed for a point pattern dataset
`X` by typing `K <- Kest(X)`.

```{r, echo=FALSE, fig.height=3, fig.width=9.5, result.width='100%'}
Klist <- anylapply(Xlist, Kest)
plot(Klist, main="")
```

### pair correlation function

The pair correlation function $g(r)$ can be defined as
$g(r) = K^\prime(r)/(2\pi r)$ where $K^\prime(r)$ is the derivative
of the $K$ function. The pair correlation function
can be interpreted as the probability that two points in the point process
will be separated by a distance equal to $r$, normalised by the corresponding
probability for a completely random (Poisson) process.

For a completely random (homogeneous
Poisson) process, $g(r) = 1$. An inhibited process will usually
have $g(r) < 1$, while a clustered process will have
$g(r) > 1$, for appropriate values of $r$.

An estimate of the pair correlation function
can be computed for a point pattern dataset
`X` by typing `g <- pcf(X)`.

```{r, echo=FALSE, fig.height=3, fig.width=9.5, result.width='100%'}
glist <- anylapply(Xlist, pcf)
plot(glist, main="")
```

## Explicit Models for clustered data

```{r}
plot(redwood)
```

### Cluster processes

A cluster process is generated in two stages.

1. a point pattern of _"parent"_ points $X$ is generated;
2. around each parent point $x_i$, a finite pattern of _"offspring"_
points $y_{i1}, \ldots, y_{in_i}$ is generated;
3. the offspring of all parents are collected together into a
single point pattern $Y$.

In a _Thomas cluster process_,

1. the parents are a homogeneous Poisson process with intensity $\kappa$;
2. each parent has a Poisson number (with mean $\mu$) of offspring,
which are displaced from the parent by independent Gaussian vectors
with standard deviation $\sigma$.

Here are simulated realisations of a Thomas process:

```{r, result.width='100%'}
plot(rThomas(kappa=10, sigma=0.2, mu=5, nsim=12),
     main="", main.panel="")
```

Maximum likelihood fitting of cluster processes is difficult because
the likelihood is quite complicated. However, 
the $K$-function of such cluster processes is known analytically,
so the model can be fitted by the method of moments (matching the
model's theoretical $K$-function to the empirical $K$-function of the
data). This is performed by the `spatstat` function `kppm`.

```{r}
fitT <- kppm(redwood ~ 1, "Thomas")
fitT
```

```{r}
plot(simulate(fitT, nsim=12))
```

```{r}
kppm(redwood ~ x+y, "Thomas")
```

### Cox processes

A _Cox process_ is formed in two steps:

1. a random function $\Lambda(u)$ is generated;
2. Given the realisation of the random function,
a Poisson point process is generated with intensity function $\Lambda(u)$.

In a _log-Gaussian Cox process_, the random function $\Lambda(u)$
is such that $\log \Lambda(u)$ is a Gaussian random function.

These models can be fitted by the same technique:

```{r}
kppm(redwood ~ x+y, "LGCP")
```

## Models for inhibited data

```{r}
plot(cells)
```

```{r}
jp <- residualspaper$Fig1
plot(jp)
```

### Gibbs models

Gibbs models were developed in theoretical physics to describe the
behaviour of molecular gases. A point pattern $x$ represents a
spatial configuration of molecules.
The probability of a particular configuration $x$ is
$$
    p(x) = Z \exp(- U(x))
$$
where $U(x)$ is the potential energy of the configuration,
and $Z$ is a normalising constant. In fact $p(x)$ is a probability density
relative to the completely random (homogeneous Poisson) point process.

To visualise this, imagine that we first generate an infinite "ensemble"
of realisations of the homogeneous Poisson process. Then each realisation
is either deleted or retained (in its entirety)
depending on its potential energy;
a realisation $x$ is retained with probability $\exp(-U(x))$.
Then what remains is an ensemble of realisations of the Gibbs process.

The simplest example is the _hard core process_ in which the points represent
the centres of discs of diameter $d$ which cannot overlap. A realisation $x$
has potential energy $U(x) = -\infty$ if any pair of points in $x$ lies closer
than distance $d$; otherwise it has potential $U(x) = 0$.
Now generate an infinite ensemble of realisations of the Poisson process.
Then delete any configuration which contains a pair of points
closer than distance $d$. The remaining realisations are an ensemble
of realisations of the hard core process.

Gibbs models can be fitted to point pattern data
by maximising Besag's _pseudolikelihood_. This is performed by `ppm`.

```{r}
ppm(cells~ 1, Hardcore())
```

```{r}
minnndist(cells)
```

```{r}
ppm(cells ~ 1, Strauss(0.1))
```

```{r}
fit <- ppm(cells ~ 1, Strauss(0.1))
plot(simulate(fit, nsim=4))
```


```{r}
plot(pcfinhom(jp))
```

```{r}
minnndist(jp)
```

```{r}
ppm(jp ~ x+y, Strauss(0.2))
```

```{r}
ppm(jp ~ x+y, Strauss(0.5))
```

### Determinantal point process models

Determinantal point process models are a relatively new class of models
with many attractive theoretical properties, but are not yet well understood
for modelling purposes.

The likelihood is determined by
the first and second moments (the intensity and the pair correlation function).
These models can be fitted in `spatstat` using the function `dppm`.

```{r}
dppm(jp ~ x+y, dppGauss)
```

```{r}
fit <- dppm(jp ~ polynom(x,y,2),
            dppMatern,
	    statistic="pcf", statargs=list(stoyan=0.2))
plot(predict(fit))	    
```

```{r}
plot(simulate(fit))
```
